---
title: Multilevel Composite Reliability
author: marklai
date: '2022-11-13'
slug: multilevel-composite-reliability
categories:
  - Statistics
tags:
  - Reliability
  - Multilevel
  - SEM
subtitle: ''
summary: ''
authors: []
lastmod: '2022-11-13T20:56:38-08:00'
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

This is a short blog post on computing multilevel reliability based on [Lai (2021)](/publication/lai2020pm).

## Load Packages

```{r}
library(lavaan)
library(semTools)
```

First, use demo data from `lavaan`

```{r}
data("Demo.twolevel", package = "lavaan")
```

Then source a script I wrote [here](https://github.com/marklhc/mcfa_reliability_supp/blob/master/multilevel_alpha.R) implementing the procedures in the paper, which defines a function `multilevel_alpha()`:

```{r}
source("https://github.com/marklhc/mcfa_reliability_supp/raw/master/multilevel_alpha.R")
```

Example syntax in R and Mplus for the paper can be found in https://github.com/marklhc/mcfa_reliability_supp. Going back to the multilevel demo data, we can consider the reliability for between-level and within-level composite scores when summing `y1`, `y2`, and `y3`. This can be done using

```{r}
multilevel_alpha(Demo.twolevel[c("y1", "y2", "y3")],
                 id = Demo.twolevel$cluster)
```

Note that the $\omega$ indices assume unidimensionality, local independence, and cross-level invariance. There were a total of six indices: three $\alpha$s and three $\omega$s, corresponding to raw composite ("2l"), between-level composite ("b"), and within-level composite ("w"). The 95% CIs were obtained using the Monte Carlo method by simulating from the asymptotic distribution of the model parameters. In addition, the function also implemented a parallel analysis procedure for determining the number of factors/components at each level, as discussed in [this paper](https://www.tandfonline.com/doi/abs/10.1080/10705511.2021.1916394). 

## Using `semTools::compRelSEM()`

Some of the above procedures were implemented in the `semTools` package. To use that, one needs to fit a two-level CFA in `lavaan` first (see `?compRelSEM`:

```{r}
mcfa_mod <- '
  level: 1
    f =~ y1 + L2*y2 + L3*y3
  level: 2
    f =~ y1 + L2*y2 + L3*y3
'
mcfa_fit <- cfa(mcfa_mod, data = Demo.twolevel, cluster = "cluster")
```

Then call the function

```{r}
compRelSEM(mcfa_fit, config = c("f"), shared = c("f"))
```


