---
title: Nominal Regression in STAN
author: marklai
date: '2022-07-30'
slug: multinomial-regression-in-stan
toc: true
categories:
  - Statistics
tags:
  - Bayesian
subtitle: ''
summary: ''
authors: []
lastmod: '2022-07-30T10:23:22-07:00'
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

## Load Packages

```{r load-pkg, message = FALSE}
library(cmdstanr)
library(dplyr)
library(ggplot2)
```

Check out this paper: https://journals.sagepub.com/doi/full/10.1177/2515245918823199

```{r stemcell}
stemcell <- read.csv("https://osf.io/vxw73/download")
```

```{r plot-stemcell}
stemcell |>
    ggplot(aes(x = rating)) +
    geom_bar() +
    facet_wrap(~ gender)
```

https://www.thearda.com/archive/files/Codebooks/GSS2006_CB.asp

The outcome is attitude towards stem cells research, and the predictor is gender.

> Recently, there has been controversy over whether the government should provide any funds at all for scientific research that uses stem cells taken from human embryos. Would you say the government . . .

- 1 = Definitely, should fund such research
- 2 = Probably should fund such research
- 3 = Probably should not fund such research
- 4 = Definitely should not fund such research

## Nominal Logistic Regression

Ordinal regression is a special case of nominal regression with the proportional odds assumption.

## Model

\begin{align}
  \text{rating}_i & \sim \mathrm{Categorical}(\pi^1_{i}, \pi^2_{i}, \pi^3_{i}, \pi^4_{i})  \\
  \pi^1_{i} & = \frac{1}{\exp(\eta^2_{i}) + \exp(\eta^3_{i}) + \exp(\eta^4_{i}) + 1}  \\
  \pi^2_{i} & = \frac{\exp(\eta^2_{i})}{\exp(\eta^2_{i}) + \exp(\eta^3_{i}) + \exp(\eta^4_{i}) + 1}  \\
  \pi^3_{i} & = \frac{\exp(\eta^3_{i})}{\exp(\eta^2_{i}) + \exp(\eta^3_{i}) + \exp(\eta^4_{i}) + 1}  \\
  \pi^4_{i} & = \frac{\exp(\eta^4_{i})}{\exp(\eta^2_{i}) + \exp(\eta^3_{i}) + \exp(\eta^4_{i}) + 1}  \\
  \eta^2_{i} & = \beta^2_{0} + \beta^2_{1} \text{male}_{i}  \\
  \eta^3_{i} & = \beta^3_{0} + \beta^3_{1} \text{male}_{i} \\
  \eta^4_{i} & = \beta^4_{0} + \beta^4_{1} \text{male}_{i} \\
\end{align}

```{r}
mod <- cmdstan_model("nominal_reg.stan")
mod
```

```{r data}
stan_dat <- with(stemcell,
     list(K = n_distinct(rating),
          N = length(rating),
          D = 1,
          y = rating,
          x = matrix(as.integer(gender == "male")))
)
```

```{r fit, message = FALSE}
# Draw samples
fit <- mod$sample(data = stan_dat, seed = 123, chains = 4, 
                  parallel_chains = 2, refresh = 500,
                  iter_sampling = 2000, iter_warmup = 2000)
```

```{r summary-fit}
fit$summary() |>
    knitr::kable()
```
