---
title: Homework 7
date: '2020-09-28'
slug: homework-7
categories: []
tags: []
linktitle: Homework 7
summary: ~
lastmod: '`r Sys.Date()`'
type: docs
output:
  blogdown::html_page:
    df_print: paged
menu:
  psyc575:
    parent: Week 9
    weight: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 100)
comma <- function(x) format(x, digits = 2, big.mark = ",")
```

## Learning Objective

- Independently analyze real data using statistical software for multilevel modeling

## Assignment Description 

In this assignment, you will use the data from the brief report by [Raykos et al. (2018)](http://dx.doi.org/10.1037/ccp0000319), which included 134 patients receiving an enhanced cognitive–behavioral therapy from public eating disorders service. All patients met the DSM-5 criteria for anorexia nervosa (AN). You will fit growth models on the eating disorder symptoms (on a range of 0 to 6) over five weekly intervals throughout treatment. Then, you will choose from a list of covariates to predict differences in the symptom trajectories. 

## Steps

```{r load-pkg, message=FALSE, include=FALSE}
# To install a package, run the following ONCE (and only once on your computer)
# install.packages("psych")  
library(here)  # makes reading data more consistent
library(tidyverse)  # for data manipulation and plotting
library(haven)  # for importing SPSS/SAS/Stata data
library(brms)  # for multilevel analysis
library(broom.mixed)  # for summarizing results
library(modelsummary)  # for making tables
# Add the following so that the LOO will be included in the msummary table
msummary_mixed <- function(models, coef_map = NULL, add_rows = NULL, ...) {
  if (is.null(coef_map)) {
    if (!"list" %in% class(models)) {
      models <- list(models)
    }
    for (model in models) {
      coef_map <- union(coef_map, tidy(model)$term)
    }
  }
  ranef_index <- grep("^(sd|cor)__", x = coef_map)
  coef_map <- c(coef_map[-ranef_index], coef_map[ranef_index])
  names(coef_map) <- coef_map
  rows <- data.frame(term = c("Fixed Effects", "Random Effects"))
  rows <- cbind(rows, rbind(rep("", length(models)), 
                            rep("", length(models))))
  length_fes <- length(coef_map) - length(ranef_index)
  attr(rows, 'position') <- c(1, (length_fes + 1) * 2)
  modelsummary::msummary(models, coef_map = coef_map, add_rows = rows, ...)
}
glance_custom.brmsfit <- function(x) {
  broom.mixed::glance(x, looic = TRUE)
}
library(interactions)  # for plotting interactions
theme_set(theme_bw())  # Theme; just my personal preference
```

1. Start with one of the templates from your previous homework. 

2. Download the data from OSF: https://osf.io/6jg78/, and save it to your `data_files` folder. The data is in `rda` format, which you can import into R using the `load()` function. 
```{r import, indent='    '}
# Import R data
load(here("data_files", "dat.rda"))
# You should see an object called `dat` on the R environment, which is the
# imported data
```

```{r, include=FALSE, }
an_dat <- dat %>% 
  select(id, age, edeq_0:edeq_4) %>% 
  pivot_longer(edeq_0:edeq_4,
               names_to = c(".value", "time"), 
               names_pattern = "(edeq)_([0-4])", 
               names_transform = list(time = as.integer))
```
Here is a spaghetti plot of the data:
```{r, echo=FALSE, warning=FALSE, indent='    '}
ggplot(an_dat, aes(x = time, y = edeq)) + 
  # add lines to connect the data
  geom_line(aes(group = id), 
            position = position_dodge(0.4), alpha = 0.5) + 
  stat_summary(fun = "mean", col = "red", size = 1, 
               geom = "line")  # add a mean trajectory
```
3. The following is the data summary of a list of covariates of interest, as well as the `edeq` variables:  
```{r}
# `age`
# `ed_chronicity`: duration of AN; time since the onset of AN in months
datasummary_skim(dat %>% select(age, ed_chronicity, edeq_0:edeq_4))
# `employed_m`: whether the participant was employed
# `education_m`: 
# `sosh`: attempted suicide or self-harmed?
datasummary_skim(dat %>% select(employed_m, education_m, sosh), 
                 type = "categorical")
```

Provide a table to compare those without missing data and those with at least one missing data on the `edeq` variables on the covariates. You can use the `datasummary_balance` function:  

```{r, eval=FALSE}
# First, add in the code for computing the complete and incomplete groups

# Then run the following:
datasummary_balance(~ complete, 
                    data = dat %>% 
                      select(complete, age, ed_chronicity, edeq_0, 
                             employed_m, education_m, sosh))
```
4. Select the `id` variable, the `edeq_0` to `edeq_4` variables, and the covariates listed above, and convert the data set to long format (you may refer to [here](../rcode8/#wide-to-long)). 
5. Fit a linear growth model with random slope of time. 
    a. Show the relevant output
    b. Interpret the fixed effect coefficients, the random intercept variance, and slope variances. 
6. Choose one of the covariates, and include your chosen variable into the model as well as its interaction with the rate of change (i.e., cross-level interaction). 
    a. Interpret the main and interaction fixed effects involving your chosen covariate in the model. 
    b. Include a plot to show the interaction.

## Grading 

Your instructor will use the rubric on Slack to grade this assignment. 

## Additional Notes

You may post your questions to the Slack channel to get help from your peers and instructor. 

