---
title: Week 4
linktitle: Lecture videos (wk 4)
type: docs
date: "2020-08-27T00:00:00+01:00"
draft: false
slides: "psyc575/04_adding_level-1_predictor"
menu:
  psyc575:
    parent: Week 4
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("Well done :+1:", 
               "You are on your way to be an MLM expert :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Overview

```{r, echo=FALSE}
blogdown::shortcode("youtube", "0bIXkUTEC3U")
```

Check your learning: The Type I error inflation problem when using OLS regression for clustered data applies to

```{r echo=FALSE, results="asis"}
check_question("Level-2 predictors and most level-1 predictors", 
               options = c("Level-1 predictors", 
                           "Level-2 predictors", 
                           "Level-2 predictors and most level-1 predictors"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Ecological Fallacy

```{r, echo=FALSE}
blogdown::shortcode("youtube", "4SnA5Du4E-c")
```

Check your learning: In a ["bizarre" research finding](https://blogs.scientificamerican.com/the-curious-wavefunction/chocolate-consumption-and-nobel-prizes-a-bizarre-juxtaposition-if-there-ever-was-one/) that found a correlation between chocolate consumption and number of Nobel prize winners at the country level, which of the following is reasonable to infer?

```{r echo=FALSE, results="asis"}
check_question("That there are potential third variables that relate to both a country's chocolate consumption level and the number of Nobel prize winners", 
               options = c("That those who eat more chocolate are more likely to win a nobel prize", 
                           "That nutritional contents in chocolate is related to cognitive functioning", 
                           "That there are potential third variables that relate to both a country's chocolate consumption level and the number of Nobel prize winners"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

Check your learning: Summarize the "Big-Fish-Little-Pond Effect" in terms of how a person's own academic performance and the overall performance of the person's school on academic self-concept. 

***

## Decomposing Effects

### Between/within effects

Note: What I called "cluster-mean centering" is the same as "within-group centering" in Snijders & Bosker (2012)

```{r, echo=FALSE}
blogdown::shortcode("youtube", "6h1U6aZmohY")
```

Check your learning: Why do we need to separate a level-1 predictor into two variables in the model?

```{r echo=FALSE, results="asis"}
check_question("Because the slope of the level-1 predictor and that of its cluster means may be different", 
               options = c("Because the slope of the level-1 predictor and that of its cluster means may be different", 
                           "Because there is an interaction between the predictor and the cluster variable", 
                           "Because the intraclass correlation for any level-1 predictor will be zero"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Path diagram and equations

```{r, echo=FALSE}
blogdown::shortcode("youtube", "ZhLOtzzxIRk")
```

Thinking exercise: Based on the between-cluster level component in the path diagram and in the equations, `meanses` can predict

```{r echo=FALSE, results="asis"}
check_question("Only the school means of math achievement", 
               options = c("Both the school-level math achievement and a student's relative standing within a school", 
                           "Only a student's relative standing within a school", 
                           "Only the school means of math achievement"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

Check your learning: Based on the results shown in the video, is the school-level slope or the student-level slope larger for the association between SES and math achievement?

```{r echo=FALSE, results="asis"}
check_question("School-level", 
               options = c("School-level", 
                           "Student-level"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Interpret the between/within effects

```{r, echo=FALSE}
blogdown::shortcode("youtube", "Pet7lt7Nb68")
```

Try it yourself: Obtain the predicted `mathach` for Student B, and compare with Students A and C. 

***

## Contextual Effects

```{r, echo=FALSE}
blogdown::shortcode("youtube", "TTCE9Tuqj0Y")
```

Check your learning: The contextual effect is

```{r echo=FALSE, results="asis"}
check_question("The predicted difference in the outcome for every unit difference in the cluster mean predictor, holding constant the level-1 predictor", 
               options = c("The predicted difference in the outcome for every unit difference in the cluster mean predictor, holding constant the level-1 predictor", 
                           "The predicted difference in the outcome for every unit difference in the level-1 predictor, within a given cluster", 
                           "The between-level effect"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Random Slopes/Random Coefficients

### Developing intuition

```{r, echo=FALSE}
blogdown::shortcode("youtube", "K6Q9wbsdgXw")
```

Check your learning: In a random-coefficient model, if there are $J$ cluster, there are

```{r echo=FALSE, results="asis"}
check_question("J intercepts, J slopes", 
               options = c("one intercept, two slopes", 
                           "one intercept, J slopes", 
                           "J intercepts, one slopes", 
                           "J intercepts, J slopes"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Equations and path diagram

```{r, echo=FALSE}
blogdown::shortcode("youtube", "_jXGrTbHft4")
```

Check your learning: Which combination of $\tau_0$ and $\tau_1$ best describes the graph below?

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
ggplot(data = tibble(x = c(-1, 3), y = c(-1, 1)), 
       aes(x = x, y = y)) + 
  geom_blank() + 
  geom_abline(data = tibble(a = rnorm(30, sd = 0.1), 
                            b = rnorm(30, mean = -0.2, sd = 0.2)), 
              aes(intercept = a, slope = b))
```


```{r echo=FALSE, results="asis"}
check_question("tau_0 = 0.1, tau_1 = 0.2", 
               options = c("tau_0 = 0.1, tau_1 = 0.2", 
                           "tau_0 = 0, tau_1 = 0.2", 
                           "tau_0 = 0.1, tau_1 = 0", 
                           "tau_0 = 0, tau_1 = 0"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Interpretations

```{r, echo=FALSE}
blogdown::shortcode("youtube", "o6PDE4IEFfg")
```

Check your learning: In a random-slope model, if $\gamma_{10}$ (the average slope) = 0.2, $\tau^2_1 = 0.04$, what is the 68% plausible range for the slopes across clusters?

```{r echo=FALSE, results="asis"}
check_question("[0, 0.4]", 
               options = c("[0, 0.4]", 
                           "[-0.48, 0.88]", 
                           "[0.16, 0.24]", 
                           "[0.12, 0.28]"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Cross-Level Interaction

<span style="color:red">There is a mistake in the path diagram, in that one of the circle should be $\beta_{1j}$, not $\beta_{0j}$</span>

```{r, echo=FALSE}
blogdown::shortcode("youtube", "7RzVD02fZHw")
```

Check your learning: Conceptually, a cross-level interaction is the same as

```{r echo=FALSE, results="asis"}
check_question("Using a cluster-level predictor to predict variations in level-1 slopes", 
               options = c("An interaction between two level-1 predictors",
                           "Using a cluster-level predictor to predict variations in level-1 slopes", 
                           "Random slopes", 
                           "Adding the main effect of a cluster-level predictor"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

