---
title: Week 2
linktitle: Lecture videos (wk 2)
type: docs
date: "2019-05-05T00:00:00+01:00"
draft: false
lastmod: '`r Sys.Date()`'
slides: "psyc575/02_review_regression"
menu:
  psyc575:
    parent: Week 2
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons")`

```{r, include=FALSE}
library(checkdown)
right_ans <- "Correct :+1:"
wrong_ans <- "That is not correct. Rewatch the video if needed"
```

## Statistical Model

```{r, echo=FALSE}
blogdown::shortcode("youtube", "axg_AshO7qM")
```

Check your learning: In the example in the video, why do we need a random component?

```{r echo=FALSE, results="asis"}
check_question("Because people who spend the same amount of time studying do not always have the same mastery level", 
               options = c("Because there is only one predictor", 
                           "Because the relation between the variables is linear", 
                           "Because people who spend the same amount of time studying do not always have the same mastery level"), 
               type = "radio", 
               right = right_ans, 
               wrong = wrong_ans)
```

***

## Import Data

```{r, echo=FALSE}
blogdown::shortcode("youtube", "NgjdsKxLMbI")
```

Check your learning: What is the coding for the `sex` variable?

```{r echo=FALSE, results="asis"}
check_question("0 = male, 1 = female", 
               options = c("1 = male, 2 = female", 
                           "2 = male, 1 = female", 
                           "0 = female, 1 = male", 
                           "0 = male, 1 = female"), 
               type = "radio", 
               right = right_ans, 
               wrong = wrong_ans)
```

```{r, echo=FALSE}
blogdown::shortcode("youtube", "UfFhGF3mJGk")
```

Take a pause and look at the scatterplot matrix. Ask yourself the following:

- How does the distribution of `salary` look?
- Are there more males or females in the data?
- How would you describe the relationship between number of publications and salary?

***

## Linear Regression

```{r, echo=FALSE}
blogdown::shortcode("youtube", "JcmmFSruAZA")
```

### Sample regression line

```{r, echo=FALSE}
blogdown::shortcode("youtube", "GIRtlYGVOXk")
```

Check your learning: How would you translate the regression line $y = \beta_0 + \beta_1 \texttt{predictor1}$ into R?

```{r echo=FALSE, results="asis"}
check_question("y ~ predictor1", 
               options = c("y ~ 0 + predictor1", 
                           "predictor1 ~ 1 + y", 
                           "y ~ predictor1"), 
               type = "radio", 
               right = right_ans, 
               wrong = wrong_ans)
```

### Centering

```{r, echo=FALSE}
blogdown::shortcode("youtube", "s19aoHq53zw")
```

Check your learning: The mean of the `pub` variable is 18.2. If we call the mean-centered version of it as `pub_c`, what should be the value of `pub_c` for someone with 10 publications?

```{r echo=FALSE, results="asis"}
check_question(" -8.2", 
               options = c(" 8.2", " 1.8", " -8.2", " 1.82"), 
               type = "radio", 
               right = right_ans, 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

## Categorical Predictor

```{r, echo=FALSE}
blogdown::shortcode("youtube", "k2nNMfJ2Xn0")
```

Check your learning: In a regression analysis, assume that there is a binary predictor that indicates whether a school is public (coded as 0) or private (coded as 1). If the coefficient for that predictor is 1.5, which category has a higher predicted score?

```{r echo=FALSE, results="asis"}
check_question("private", 
               options = c("private", "public"), 
               type = "radio", 
               right = right_ans, 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

## Multiple Regression

```{r, echo=FALSE}
blogdown::shortcode("youtube", "O2ob6wC91w0")
```

Think more: the coefficient of `pub_c` becomes smaller after adding `time` into the equation. Why do you think that is the case?

***

## Interaction

```{r, echo=FALSE}
blogdown::shortcode("youtube", "cqGZD8s5WvU")
```

```{r, echo=FALSE}
blogdown::shortcode("youtube", "vK-2xMftORM")
```

Pratice yourself: from the interaction model obtain the regression line when `pub` = 50. 
