---
title: Week 6
linktitle: Lecture videos (wk 6)
type: docs
date: "2020-09-10T00:00:00+01:00"
draft: false
slides: "psyc575/07_mlm_for_experimental_data"
menu:
  psyc575:
    parent: Week 6
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("You got it :+1:", 
               "Great job :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Multilevel Experiments

```{r, echo=FALSE}
blogdown::shortcode("youtube", "9btZ_FrDGus")
```

***

Check your learning: In a research study, 10 hospitals are randomly assigned to a treatment condition to adopt a new drug, whereas the other 10 hospitals use the conventional method. What is the design of this study?

```{r echo=FALSE, results="asis"}
check_question("A cluster-randomized trial", 
               options = c("A cluster-randomized trial", 
                           "A multisite trial", 
                           "A within-subjects design"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Example

```{r, echo=FALSE}
blogdown::shortcode("youtube", "4baFBogxmAs")
```

***

Check your learning: In the data set, how many observations are there at level 1?

```{r echo=FALSE, results="asis"}
check_question("7,803", 
               options = c("7,803", 
                           "153", 
                           "51"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

## Long vs. Wide data set

```{r, echo=FALSE}
blogdown::shortcode("youtube", "NsRnhNz3-10")
```

***

Check your learning: In the following data, `hvltt` to `hvltt4` are the test score of a verbal learning test across four time points. Is this a long or a wide data set?

```{r, message=FALSE}
library(tidyverse)
read.csv("https://books.psychstat.org/rdata/data/active.csv",
         header = TRUE) %>% 
  select(id, hvltt:hvltt4)
```


```{r echo=FALSE, results="asis"}
check_question("wide", 
               options = c("long", 
                           "wide"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans, 
               alignment = TRUE)
```

## Crossed Random Levels

```{r, echo=FALSE}
blogdown::shortcode("youtube", "Vgv78TgDWgA")
```

***

Think more: What is the data structure if there are 1,000 students from 100 schools and 30 neighborhoods, and each school has students from multiple neighborhoods?

***

### Unconditional model

```{r, echo=FALSE}
blogdown::shortcode("youtube", "zECzzA4H_uo")
```

***

Practice yourself: Compute the design effects for the person level and for the item level. Do the design effects suggest the need for both levels? (That is, are both design effects > 1.1?)

***

## Cross-Classified Random-Effect Model With Random Slopes

```{r, echo=FALSE}
blogdown::shortcode("youtube", "dyDKHC0sugE")
```

***

Check your learning: If in the experiment, each person respond to each item 3 times, each time with a different `duration`. At what level(s) can `duration` have random slopes?

```{r echo=FALSE, results="asis"}
check_question("Both person and item", 
               options = c("Both person and item", 
                           "Person only", 
                           "Item only", 
                           "Within-cell"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Full Model

```{r, echo=FALSE}
blogdown::shortcode("youtube", "bIrEtjB91gk")
```

## Effect Size

Please check out the slides and the examples in the [R code](../rcode7/#effect-size)

