---
title: Week 13
linktitle: Lecture videos (wk 13)
type: docs
date: "2020-10-31T13:54:00+01:00"
draft: false
slides: "psyc575/13_samplesize_planning"
menu:
  psyc575:
    parent: Week 13
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("Correct :+1:", 
               "Great job :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Sample Size Planning

```{r, echo=FALSE}
blogdown::shortcode("youtube", "1NLFAQ8M4Ak")
```

## Precision Analysis

### Sample size and standard error

```{r, echo=FALSE}
blogdown::shortcode("youtube", "Bors1lcfInA")
```

### Example

```{r, echo=FALSE}
blogdown::shortcode("youtube", "wqrlAXWLC2I")
```

***

Practice yourself: Find the required $J$ for *SE*($\gamma_{10}$) to be $\leq .1$, when the treatment variable is at level-1 (i.e., $X$), with 10 individuals per cluster, ICC = $\tau^2_0$ = 0.3, $\sigma^2$ = 0.7, and $\tau^2_1 = 0.1$. 

***

## Power Analysis

```{r, echo=FALSE}
blogdown::shortcode("youtube", "7nYrIV5nbDs")
```

***

Check your learning: In the examples in the video, what would change if the researcher assumes a larger effect under the alternative hypothesis?

```{r echo=FALSE, results="asis"}
check_question("power", 
               options = c("power", 
                           "critical region", 
                           "Type I error rate"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Tools

```{r, echo=FALSE}
blogdown::shortcode("youtube", "clV_vBac6cM")
```

#### PowerUpR Demo

```{r, echo=FALSE}
blogdown::shortcode("youtube", "d9dB9m6wptM")
```

***

Practice yourself: Using PwerUpR, verify that power = .678 when the treatment variable is at level 1 (i.e., Two-level MS-IRT, random treatment), with an average effect of 0.3 (standardized), no covariate, ICC = 0.3, $\tau^2_1 = 0.15$ (i.e., `omega2` = 0.5 in the program), with 30 clusters and 10 observations per cluster. 

Now, include 1 covariate that has an $R^2$ of .40. What is the power?

***

### Notes on power analysis

```{r, echo=FALSE}
blogdown::shortcode("youtube", "n73ZFKKlUQw")
```

***

Check your learning: A researcher originally plan to collect data from 30 classrooms with 5 students per classroom, but then found out the study was underpowered. He can collect data from 150 more students. Would it be more beneficial to make it 300 students from 30 classrooms or 300 students from 60 classrooms? Why?

```{r echo=FALSE, results="asis"}
check_question("300 students from 60 classrooms, because the larger sample size for the classroom level gives more power for level-2 coefficients", 
               options = c("300 students from 30 classrooms, because it makes the cluster size larger", 
                           "300 students from 30 classrooms, because more classrooms increase the noise in the data", 
                           "300 students from 60 classrooms, because the larger sample size for the classroom level makes tau^2_0 smaller", 
                           "300 students from 60 classrooms, because the larger sample size for the classroom level gives more power for level-2 coefficients"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***
