---
title: Week 11
linktitle: Lecture videos (wk 11)
type: docs
date: "2020-10-03T00:00:00+01:00"
draft: false
slides: "psyc575/11_prediction"
menu:
  psyc575:
    parent: Week 11
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("Correct :+1:", 
               "Great job :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Predictive Models in MLM

```{r, echo=FALSE}
blogdown::shortcode("youtube", "yq6fGgU7UiI")
```

```{r, echo=FALSE}
blogdown::shortcode("youtube", "6YyRwbEhzqE")
```

***

Think more: Think about a prominent theory in your area of research. What predictions does it make? Does it give precise predictions? 

***

## Example

```{r, echo=FALSE}
blogdown::shortcode("youtube", "jCP1OHs504M")
```

***

Check your learning: In a multilevel model with students nested within schools and with student math achievement as the outcome variable, what is a cluster-specific prediction?

```{r echo=FALSE, results="asis"}
check_question("The predicted achievement for a student from one of the schools included in the sample.", 
               options = c("The predicted achievement for a student from one of the schools included in the sample.", 
                           "The predicted mean achievement of a new school.", 
                           "The predicted achievement for a student from a new school that are not part of the sample."), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Prediction Error

```{r, echo=FALSE}
blogdown::shortcode("youtube", "CCRqTspEXlc")
```

## Overfitting

```{r, echo=FALSE}
blogdown::shortcode("youtube", "QXjW5vZV--E")
```

***

Check your learning: Which of the following growth curve model would show the largest degree of overfitting, given a sample of 15 participants across 5 time points?

```{r echo=FALSE, results="asis"}
check_question("A piecewise growth model with four phases.", 
               options = c("A linear growth model.", 
                           "A piecewise growth model with two phases.", 
                           "A piecewise growth model with four phases."), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Out-Of-Sample Prediction Errors

```{r, echo=FALSE}
blogdown::shortcode("youtube", "vVQpY1vh208")
```

***

Check your learning: Why shouldn't we just choose a model with the lowest in-sample prediction error?

```{r echo=FALSE, results="asis"}
check_question("A model that fits one data set too well usually does not fit well with other data sets.", 
               options = c("Prediction error should not be a criterion for choosing among models.", 
                           "One should use a likelihood ratio test instead.", 
                           "A model that fits one data set too well usually does not fit well with other data sets."), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Cross Validation

```{r, echo=FALSE}
blogdown::shortcode("youtube", "8NoYZUzGfGQ")
```

***

Check your learning: Why does cross-validation give a better estimate of the out-of-sample prediction error?

```{r echo=FALSE, results="asis"}
check_question("It uses different subsets of data for estimation and for prediction.", 
               options = c("It uses different subsets of data for estimation and for prediction.", 
                           "It is developed in the machine learning literature.", 
                           "It uses less data for estimation, so the results are more stable."), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Information Criterion

```{r, echo=FALSE}
blogdown::shortcode("youtube", "rxRGk1m1Pvo")
```

## Regularization

```{r, echo=FALSE}
blogdown::shortcode("youtube", "WQM6uNbCVCk")
```

***

Check your learning: Under which of the following situations is the model most prone to overfitting (and thus would benefit most from regularization)?

```{r echo=FALSE, results="asis"}
check_question("N = 100, p_loo = 100", 
               options = c("N = 100, p_loo = 100", 
                           "N = 1000, p_loo = 10", 
                           "N = 10000, p_loo = 100"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***
