---
title: Week 10
linktitle: Lecture videos (wk 10)
type: docs
date: "2020-10-03T00:00:00+01:00"
draft: false
slides: "psyc575/10_longitudinal_analysis_2"
menu:
  psyc575:
    parent: Week 10
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("Correct :+1:", 
               "Great job :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Longitudinal Data Analysis II

```{r, echo=FALSE}
blogdown::shortcode("youtube", "xW_84TYK85g")
```

```{r, echo=FALSE}
blogdown::shortcode("youtube", "F-NNPPAuPWE")
```

***

Check your learning: When analyzing a conversation between a couple, a researcher is interested in whether a person follow up the partner's complaints with positive or negative behaviors. Is this an example of studying trends or fluctuations?

```{r echo=FALSE, results="asis"}
check_question("Fluctuations", 
               options = c("Trends", 
                           "Fluctuations"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

## Model 1

```{r, echo=FALSE}
blogdown::shortcode("youtube", "mDdyowlEUic")
```

***

Check your learning: In the model above, what is the interpretation of the contextual effect of `mood1`?

```{r echo=FALSE, results="asis"}
check_question("The predicted difference in symptoms between two individuals with 1 unit difference in average negative mood on a day when both individuals have the same daily mood level", 
               options = c("The predicted difference in average symptoms between two individuals with 1 unit difference in average negative mood", 
                           "The predicted difference in symptoms for an individual across two days with 1 unit difference in mood", 
                           "The predicted difference in symptoms between two individuals with 1 unit difference in average negative mood on a day when both individuals have the same daily mood level"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Model 2

```{r, echo=FALSE}
blogdown::shortcode("youtube", "bbNaNi2hUXQ")
```

Note: For the coefficients of `stressor` and `stressor_pm` in the above model, the coefficients are ones adjusting for the other predictors in the model (e.g., `mood1_pm`, `mood1_pmc`, `women`, and their interactions). 
