---
title: Week 7
linktitle: Lecture videos (wk 7)
type: docs
date: "2020-09-10T00:00:00+01:00"
draft: false
slides: "psyc575/06_Bayesian_mlm"
menu:
  psyc575:
    parent: Week 7
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("You got it :+1:", 
               "Your statistical knowledge is getting better and better :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Bayesian methods

```{r, echo=FALSE}
blogdown::shortcode("youtube", "NqDqaNKf6DE")
```

***

Check your learning: What does MCMC stand for?

```{r echo=FALSE, results="asis"}
check_question("Markov Chain Monte Carlo", 
               options = c("Multiple-Chain Monte Carlo", 
                           "Markov Chain Monte Carlo"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Bayesian vs. Frequentist Probabilities

```{r, echo=FALSE}
blogdown::shortcode("youtube", "UxRi9sakqmk")
```

***

Check your learning: Which of the following is allowed with Bayesian but not with Frequentist interpretations of probability?

```{r echo=FALSE, results="asis"}
check_question("Assigning a probability to a model parameter", 
               options = c("Assigning a probability to a model parameter", 
                           "Computing the probability of the data", 
                           "Use probability to describe the distribution of fixed-effect estimates"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Why Bayesian?

```{r, echo=FALSE}
blogdown::shortcode("youtube", "huC96JA_OPY")
```

## The Prior Distribution

```{r, echo=FALSE}
blogdown::shortcode("youtube", "augKnNQIRYo")
```

***

Check your learning: In a model predicting happiness from income, how do you understand the prior distribution for the slope coefficient?

```{r echo=FALSE, results="asis"}
check_question("The probability distribution that describes the prior belief of what the slope should be, without information from the data", 
               options = c("The sampling distribution of the estimate across repeated samples", 
                           "The probability distribution that describes our uncertainty in the data", 
                           "The probability distribution that describes the prior belief of what the slope should be, without information from the data"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Bayes's Theorem

```{r, echo=FALSE}
blogdown::shortcode("youtube", "jWfrTfsJYq4")
```

***

Check your learning: Based on the Bayes's theorem, what is $P(\gamma | y)$?

```{r echo=FALSE, results="asis"}
check_question("The posterior distribution of gamma, incorporating the information from data y", 
               options = c("The likelihood function of gamma", 
                           "The prior distribution of gamma", 
                           "The posterior distribution of gamma, incorporating the information from data y"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## The Posterior Distribution

```{r, echo=FALSE}
blogdown::shortcode("youtube", "4i5ER_Qrgus")
```

***

Check your learning: The Bayes estimation of $\gamma_{01}$ is

```{r echo=FALSE, results="asis"}
check_question("The posterior distribution of gamma, incorporating the information from data y", 
               options = c("The median value in the posterior distribution of gamma_01", 
                           "The mean value in the prior distribution of gamma_01", 
                           "The mean value in the posterior distribution of gamma_01", 
                           "The maximum value in the likelihood function of gamma_01"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Markov Chain Monte Carlo

```{r, echo=FALSE}
blogdown::shortcode("youtube", "6Wg7iLR8IDE")
```

```{r, echo=FALSE}
blogdown::shortcode("youtube", "4VEVyPz_62w")
```

## Handling Assumption Violations

Please check out examples in [R code](../rcode6/#using-brms-to-relax-assumptions)

