---
title: Week 9
linktitle: Lecture videos (wk 9)
type: docs
date: "2020-10-03T00:00:00+01:00"
draft: false
slides: "psyc575/09_longitudinal_analysis_1"
menu:
  psyc575:
    parent: Week 9
    weight: 1
output:
  blogdown::html_page:
    toc: true
    df_print: paged
# Prev/next pager order (if `docs_section_pager` enabled in `params.toml`)
weight: 1
---

`r blogdown::shortcode("slide-buttons2")`

```{r, include=FALSE}
library(checkdown)
right_ans <- c("Correct :+1:", 
               "Great job :+1:")
wrong_ans <- "Try again. You can do it!"
```

## Longitudinal Data Analysis

```{r, echo=FALSE}
blogdown::shortcode("youtube", "bdDrcmktBnE")
```

***

Check your learning: In a research study, data were collected for a group of patients on symptoms of eating disorder on a weekly interval across 5 weeks. What type of data is this?

```{r echo=FALSE, results="asis"}
check_question("Panel data", 
               options = c("Panel data", 
                           "Intensive longitudinal data", 
                           "Cross-sectional data"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Example data

```{r, echo=FALSE}
blogdown::shortcode("youtube", "3PDA_ur1sOU")
```

***

Check your learning: In the data set, at what level is `homecog`, which is a measure of mother's cognitive stimulation at baseline?

```{r echo=FALSE, results="asis"}
check_question("Level 2", 
               options = c("Level 1", 
                           "Level 2"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

### Basic attrition analysis

```{r, echo=FALSE}
blogdown::shortcode("youtube", "b2dto5JwRQ8")
```

See the [R code section](../rcode8/#attrition-analyses). 

***

Check your learning: In the spaghetti plot, what does the average trend line mean?

```{r echo=FALSE, results="asis"}
check_question("That the sample average reading score increased across time", 
               options = c("That every child showed an increasing trend", 
                           "That the sample average reading score increased across time", 
                           "That there is no variability in individual trajectories"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

## Growth Curve Modeling

```{r, echo=FALSE}
blogdown::shortcode("youtube", "iT7kzf2gIUs")
```

***

Thinking exercise: In a growth model, what does it mean when $\tau_1 = 0$?

```{r echo=FALSE, results="asis"}
check_question("That every participant has the same rate of change", 
               options = c("That on average, there is no change on the outcome across time", 
                           "That every participant has the same rate of change", 
                           "That every participant has the same outcome score"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Linear growth

```{r, echo=FALSE}
blogdown::shortcode("youtube", "4xHz4OIPVh8")
```

***

Check you learning: What is the advantage of having time to start at 0?

```{r echo=FALSE, results="asis"}
check_question("The intercept corresponds to the predicted score at the first time point", 
               options = c("The intercept corresponds to the predicted score at the first time point", 
                           "So that tau_1 will be the SD of the slopes"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***

### Piecewise linear growth

```{r, echo=FALSE}
blogdown::shortcode("youtube", "p6TQdUzi1d0")
```

***

Practice yourself: What should the coding of phase 1 and phase 2 be if the turning point is set at `time` = 2?

***

```{r, echo=FALSE}
blogdown::shortcode("youtube", "MbDcv05wOrU")
```

Note: In this example, the turning point was chosen mostly based on the spaghetti plot and was arbitrary. For your research, you should justify your choice. 

***

Check your learning: If a piecewise growth model has an LOOIC of 23745, and a linear growth model has an LOOIC of 23650, which model should be preferred?

```{r echo=FALSE, results="asis"}
check_question("Linear", 
               options = c("Piecewise", 
                           "Linear"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans, 
               alignment = TRUE)
```

***

## Time-Invariant Covariates 

```{r, echo=FALSE}
blogdown::shortcode("youtube", "fI6XJnVQ3pg")
```

***

Thinking exercise: What does the coefficient for `phase1` mean when the model includes an interaction between `phase1` and `homecog9`?

```{r echo=FALSE, results="asis"}
check_question("The average slope from time 0 to time 1 for kids with homecog = 9", 
               options = c("The average slope from time 0 to time 1", 
                           "The average slope from time 0 to time 1 for kids with homecog = 9", 
                           "The average slope from time 0 to time 1, holding constant homecog"), 
               type = "radio", 
               right = sample(right_ans, 1, prob = c(.7, .3)), 
               wrong = wrong_ans)
```

***
