---
title: 'Exercise: Interpreting interactions'
date: "8/20/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-pkg, message=FALSE, include=FALSE}
library(here)  # makes reading data more consistent
library(tidyverse)  # for data manipulation and plotting
library(interactions)  # for plotting interactions
theme_set(theme_bw())  # Theme; just my personal preference
```

```{r salary_dat, results='hide'}
# The `here()` function forces the use of the project directory
here("data_files", "salary.txt")
# Read in the data
salary_dat <- read.table(here("data_files", "salary.txt"), header = TRUE)
```

The sample means of `time` and `pub` are:

```{r mean-time-pub}
mean(salary_dat$time)
mean(salary_dat$pub)
```


```{r m4}
salary_dat <- salary_dat %>% 
  mutate(pub_c = pub - mean(pub), 
         time_c = time - mean(time))
# Fit the model with interactions:
m4 <- lm(salary ~ pub_c * time_c, data = salary_dat)
summary(m4)  # summary
```

Here's the code and the plot in the note:

```{r interaction-plot}
interactions::interact_plot(m4,
                            pred = "pub_c",
                            modx = "time_c",
                            # Insert specific values to plot the slopes. 
                            # Pay attention that `time_c` has been centered
                            modx.values = c(1, 7, 15) - 6.79, 
                            modx.labels = c(1, 7, 15), 
                            plot.points = TRUE, 
                            x.label = "Number of publications (mean-centered)", 
                            y.label = "Salary", 
                            legend.main = "Time since Ph.D.")
```


## Instruction

First, think about how to answer the following questions on your own, for a few minutes. Then when you get to the breakout room, select a group reporter, and continue to work on the following questions: 

1. Write down the regression equation for the above model. 

2. Professor A has graduated 6.79 years since Ph.D. What should be their value of `time_c`?

3. Using the regression equation, compute the predicted salary for the following four faculty
    a. Professor A: 6.79 years after Ph.D. (i.e., `time_c` = ?), 18.18 publications (i.e., `pub_c` = ?)
    b. Professor B: 7.79 years after Ph.D., 18.18 publications
    c. Professor C: 6.79 years after Ph.D., 28.18 publications
    d. Professor D: 7.79 years after Ph.D., 28.18 publications
4. Professors A and B differ by 1 year of experience. What is the difference in predicted salary between them? Does that correspond to any of the model coefficients?
5. Professors C and D also differ by 1 year of experience. What is the difference in predicted salary between them? Is that difference the same as in (4)?
6. Provide an interaction plot with time since Ph.D. on the x-axis and salary on the y-axis. 

