---
title: 'Exercise: Empirical Bayes Estimates'
date: "9/03/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-pkg, message=FALSE, include=FALSE}
library(here)  # makes reading data more consistent
library(tidyverse)  # for data manipulation and plotting
library(haven)  # for plotting interactions
library(lme4)  # for running a mixed effect model
theme_set(theme_bw())  # Theme; just my personal preference
```

```{r import_sav, message=FALSE}
# Read in the data (pay attention to the directory)
hsball <- read_sav(here("data_files", "hsball.sav"))
```

```{r, message=FALSE}
m0 <- lmer(mathach ~ (1 | id), data = hsball)
summary(m0)
```

Consider school 8367, which has 
`r hsball %>% filter(id == "8367") %>% nrow()` students in our sample:

```{r, echo=FALSE, results='asis'}
hsball %>% 
  filter(id == "8367") %>% 
  knitr::kable()
```

The sample mean of `mathach` is:

```{r}
hsball %>% 
  filter(id == "8367") %>% 
  summarize(mean(mathach))
```

However, if we rely on the random intercept model, the estimate of $\beta_{0j}$ (i.e., the empirical Bayes estimate, $\hat \beta_{0j}$) is:

```{r}
predict(m0, newdata = data.frame(id = "8367", mathach = NA))
```

## Instruction

First, select a group reporter. Then answer the following questions: 

Questions:

1. In the random intercept model, we want to know whether there are differences in the means of math achievement across schools. What statistical method you have learned answer a similar question?

2. Why do you think there is such a difference?  

3. Which one (sample mean, 4.55 or $\hat \beta_{0j} = 6.53$) do you think is
closer to the actual mean for everyone (i.e., 153 students) in that school?

4. The formula for the empirical Bayes estimate is 
$$\hat \lambda_j \times \text{sample cluster mean} + (1 - \hat \lambda_j) \times \hat \gamma_{00}, $$
where $\hat \lambda_j = \frac{\hat \tau^2_0}{\hat \tau^2_0 + \hat \sigma^2 / n_j}$ and $n_j$ is the sample cluster size for the $j$th cluster. 
    i. Verify that $\hat \lambda_j = 0.755$ for school 8367. 
    ii. Verify that $\hat \beta_{0j} = 6.53$ for school 8367. 

5. What will be $\hat \beta_{0j}$ for school 8367 if instead there were 70 students in the sample (i.e., $n_j = 70$)? What does it say about the relation between shrinkage and cluster size?
