---
title: 'Exercise: Decomposing Effects'
date: "9/10/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-pkg, message=FALSE, include=FALSE}
library(here)  # makes reading data more consistent
library(tidyverse)  # for data manipulation and plotting
library(haven)  # for plotting interactions
library(lme4)  # for running a mixed effect model
library(broom.mixed)
library(modelsummary)
theme_set(theme_bw())  # Theme; just my personal preference
```

## Data Description: Hoffman (2014, p. 24)

The data is the first wave of the Cognition, Health, and Aging Project. 

> The Cognition, Health, and Aging Project (CHAP) consists of a sample of both younger and older adults collected using a measurement burst design. A single measurement burst included six observations over a two-week period. Bursts were then separated by 6-month intervals. In this manner, both shorter-term (within-burst) and longer-term (between-burst) change was observed during the study. A variety of measures of physical, cognitive, and emotional well-being were collected (for more information about the CHAP data, see Sliwinski, Almeida, Smyth, & Stawski, 2009; Sliwinski, Smyth, Hofer, & Stawski, 2006; and Stawski, Sliwinski, Almeida, & Smyth, 2008). 

```{r import_sav, message=FALSE}
# Download from the Internet, unzip, and read in
tmp <- tempfile()
download.file("https://www.pilesofvariance.com/Chapter8/SPSS/SPSS_Chapter8.zip", tmp)
stress_data <- read_sav(unz(tmp, "SPSS_Chapter8/SPSS_Chapter8.sav"))
stress_data
```

Here is a quick plot between negative mood (`mood`) and daily number of symptoms (`symptoms`):

```{r scatter, warning=FALSE}
ggplot(stress_data, aes(x = mood, y = symptoms)) + 
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.2) + 
  geom_smooth()
```


## Instruction

First, select a group reporter. Then answer the following questions: 

Questions:

1. Describe the hierarchical structure of the data. Which level is level 1 and which is level 2? What is the sample size ($N$ and $J$) for each level?

2. Use the `modelsummary::datasummary()` function to obtain the mean, median, sd, minimum, and maximum values of negative mood (`mood`), age at baseline (`baseage`), and number of physical symptoms (`symptoms`). Copy your R syntax to the spreadsheet
```{r descriptive, indent='    '}
# Modify the code below. You need to use `Min()` instead of `min()` as there are
# missing data
datasummary(mood ~ (Mean + SD), data = stress_data)
```

3. Run an MLM with negative mood predicting number of physical symptoms. What is the estimated fixed-effect coefficient? Does the 95% CI include zero?
```{r smushed, indent='    '}

```

4. Run an MLM that separates the association of negative mood and physical symptoms into the between-person and the within-person associations. Record the corresponding fixed-effect coefficient estimates. Are they similar? Do the 95% CIs include zero?
```{r m_bw, indent='    '}
# Cluster-mean centering (edit the code below)
stress_data <- stress_data %>% 
  group_by() %>% 
  mutate()
# Model

```
5. Use a graph (either by hand or with R) to show the between-person and within-person slopes of negative mood predicting symptoms. 

6. How would you interpret the contextual effect in this context?

