---
title: Piping with `magrittr`
author: marklai
date: '2020-12-16'
slug: piping-with-magrittr
categories:
  - Programming
tags:
  - R
subtitle: ''
summary: ''
authors: []
lastmod: '2020-12-16T14:38:37-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

I have just spent a semester teaching [multilevel modeling](/courses/psyc575), and in the R codes I provided, I usually use the [pipe operator (`%>%`)](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html). For example, to compute the cluster means, we can do

```{r Hsb82, message=FALSE}
library(tidyverse)
data("Hsb82", package = "mlmRev")
Hsb82 <- Hsb82 %>% 
  group_by(school) %>% 
  mutate(ses_cm = mean(ses)) %>% 
  ungroup()
```

However, it's kind of embarassing that I only recently found out the assignment pipe (`%<>%`) operator, as discussed [here](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html#additional-pipe-operators). For example, 

```{r x-pipe, message=FALSE}
library(magrittr)
set.seed(123)
x <- rnorm(10)
mean(x)
# Add 1 to x
x %<>% magrittr::add(1)
mean(x)
# The above is equivalent to 
# x <- x + 1
```

For the cluster mean example, we can do

```{r magrittr-assignment}
Hsb82 %<>% 
  group_by(school) %>% 
  mutate(ses_cm2 = mean(ses)) %>% 
  ungroup()
select(Hsb82, ses_cm, ses_cm2)
```

which saves the additional typing of `Hsb82 <- Hsb82 %>%`. That said, the `%<>%` is not commonly seen when reading other people's code, so perhaps the R community still prefer just using the `%>%` operator. But it's at least good to know there is a potentially more convenient way. There is also the `%$%` and `%T>%` operator, as discussed in [this vignette](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html). 
